Increasing Productivity Amid Stable Nutrient Regimes in Rhode Island Lakes and Reservoirs

### 2019-04-08 manuscript/manuscript.Rmd
still need to force length of x-axis to 1993 to 2016...

---
title: "Long-term Water Quality Change in Rhode Island Lakes and Ponds"
author:
- Hollister. J. W. ^\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Kreakie, B. J. ^*1*^, Shivers,
  S. ^*3*^, Milstead, W. Bryan. ^*1*^, Herron, E. ^*2*^, Green, L. ^*2*^, Gold, A
  ^*2*^
- ^*1*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
  Ecology Division, Narragansett, RI 02882
- ^*2*^ University of Rhode Island, Department of Natural Resources Science, Kingston,
  RI 02881
- ^*3*^ ORISE, Narragansett, RI 02882
- '^\*^  *corresponding author: hollister.jeff@epa.gov*'
bibliography: manuscript.bib
output:
  word_document:
    reference_docx: manuscript_template.docx
editor_options:
  chunk_output_type: inline
csl: plos.csl
abstract: The University of Rhode Island's Watershed Watch Volunteer Monitoring
  Program has been collecting water quality data on Rhode Island lakes and ponds for
  close to three decades, allowing us to explore long-term trends in common water
  quality parameters.  Not all lakes and ponds in the study area were sampled across
  the full time period and lakes were often added in geographic clusters (e.g. in
  urbanized northern Rhode Island).  Not unlike how long-term temperature records
  are analyzed, we centered and scaled (i.e., the z-score) water quality measurements
  on a per-station basis. This provides a robust and commonly scaled measurement to
  explore this data for long-term trends.  State-wide aggregation shows increasing
  temperature, chlorophyll *a*, and total nitrogen. Interestingly, total phosphorus
  is showing a decline, perhaps reflecting the management focus on phosphorus reductions.   While yearly trends are useful, they do mask month-to-month variability differences 
  across sites.  Additionally, while most sites track the yearly trend in decreasing water
  quality, there are bright spots with some sites improving over the 25 years. Contrary
  to previously reported analyses that show relatively stable water quality at the
  regional scale, our analysis shows that long-term water quality trends within Rhode
  Island show some parameters improving while others are in decline.  Importantly,
  this analysis also points out the incredible value and importance of data from long-term
  monitoring programs, like the Univeristy of Rhode Island's Watershed Watch, for
  identifying trends in environmental condition.
---

>>>>

---
title: "Increasing Productivity Amid Stable Nutrient Regimes in Rhode Island Lakes and Reservoirs"
author:
- Hollister. J. W. ^\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Kreakie, B. J. ^*1*^, Shivers,
  S. ^*3*^, Milstead, W. Bryan. ^*1*^, Herron, E. ^*2*^, Green, L. ^*2*^, Gold, A
  ^*2*^
- ^*1*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
  Ecology Division, Narragansett, RI 02882
- ^*2*^ University of Rhode Island, Department of Natural Resources Science, Kingston,
  RI 02881
- ^*3*^ ORISE, Narragansett, RI 02882
- '^\*^  *corresponding author: hollister.jeff@epa.gov*'
bibliography: manuscript.bib
output:
  word_document:
    reference_docx: manuscript_template.docx
editor_options:
  chunk_output_type: inline
csl: plos.csl
abstract: "The University of Rhode Island's Watershed Watch Volunteer Monitoring Program has been collecting water quality data on Rhode Island lakes and ponds for close to three decades, allowing us to explore long-term trends in common water quality parameters.  Not all lakes and ponds in the study area were sampled across the full time period and lakes were often added in geographic clusters (e.g. in urbanized northern Rhode Island).  Not unlike how long-term temperature records are analyzed, we centered and scaled (i.e., the z-score) water quality measurements on a per-station basis. This provides a robust and commonly scaled measurement to explore this data for long-term trends.  State-wide aggregation shows increasing temperature, chlorophyll *a*, and total nitrogen. Interestingly, total phosphorus is showing a decline, perhaps reflecting the management focus on phosphorus reductions. Contrary to previously reported analyses that show relatively stable water quality at the regional scale, our analysis shows that long-term water quality trends within Rhode Island show some parameters improving while others are in decline. Importantly, this analysis also points out the incredible value and importance of data from long-term monitoring programs, like the Univeristy of Rhode Island's Watershed Watch, for identifying trends in environmental condition."
---
